---
title: "Correlating water quality fluctuations with sensor data"
subtitle: "Vitens Open Data Challenge: Smart Water Grid"
author: "Dennis van den Berg"
date: "21/01/2016"
output: html_document
---


## Introduction

Central question: What are the correlations between changes in water quality measued by Eventlab sensors and other real-time measurements, statuses and alarm values?

See documentation: https://github.com/dljvandenberg/smart_water_grid/blob/master/doc/Vitens%20Data%20Challenge%2007122015.pdf


## Data Description

```{r, message=FALSE, echo=FALSE}
library(dplyr)
library(reshape2)
library(lubridate)
library(ggplot2)
library(caret)

dir.data <- "~/git/smart_water_grid/data"
setwd(dir.data)

createvariablefromfilename <- function(filename) {
    # Chop off beginning and end of filename and convert to valid variable name
    variablename <- make.names(gsub("(^Dump_|.csv$)", "", filename))
}

readfile <- function(file) {
    variablename <- createvariablefromfilename(file)
    
    data <- read.csv(file, comment.char = "#", col.names = c("Time", variablename), colClasses = c("character", "numeric"),
                     na.strings = c("", " ", "CalcFailed", "Calc Failed", "Bad", "BadInput", "Bad Input", "PtCreated", "Pt Created", "CommFail", "ScanOff", "Configure", "I/OTimeout"))
    
    # Convert to POSIXct/POSIXt time format
    data$Time <- ymd_hms(data$Time)
    
    return(data)
}

readfiles <- function(fileslist) {
    listofdataframes <- lapply(fileslist, function(file) readfile(file))
    merged <- Reduce(function(x, y) merge(x, y, all=TRUE), listofdataframes)

    return(merged)
}

selecttimerange <- function(dataframe, begintime = -Inf, endtime = Inf) {
    subset(dataframe, Time >= begintime & Time <= endtime)
}
```


### Data files

```{r, echo=FALSE}
files.all <- list.files(path=dir.data, pattern="csv$")
```

Our data set contains `r length(files.all)` files and its total size is a little over 5 GB. All files are in .csv format and contain Timestamp-Value pairs. Furthermore the data files contain some additional information in the header comments. 

The file name contains the variable that was measured, in format `Dump_<variable>.csv`:

```{r, echo=FALSE}
head(files.all, n=4)
```



### Time range

According to the documentation measurements are done in the time range January 2014 to December 2015. We confirmed this by looking at the data. We also observe that some variables are only available in specific time ranges, often data is missing in large blocks of this 2-year time range.

```{r, warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE}
setwd(dir.data)
files.select <- list.files(path=dir.data, pattern="vitnor.*csv$")[1]
data.select <- readfiles(files.select)
summary(data.select)
ggplot(data.select, aes_string(x="Time", y=names(data.select)[2])) + geom_point() + ggtitle("Raw data (unaggregated)")
```


### Variable overview

In the documentation we see the following explanation of variable names:

Quantity      | Variable code    | Remarks
------------- | -------------    | -------------
Eventlab data | `*vitnor*`       | 3 per sensor (`vitnor1`, `vitnor2`, `vitnor3`). Alarm codes are given in case 1 or more values above threshold (code Orange for `1 > max(vitnor*) >= 1.5`, code Red for `max(vitnor*) > 1.5`)
Temperature   | `*TM*`           | 
Flow          | `*FT*`, `*VO*`   | Negative values allowed (in case of opposite flow direction)
Pressure      | `*PT*`, `*DO*`   | We found that these had to be followed by 2 digits (`*PTxx*`, `*DOxx*`) in order not to select incorrect variables 
Conductivity  | `FR-PNB_TR00QI03PV*`, `FR-POH_-TR00QI03PV*` | 
Acidity (pH)  | `FR-PNB_TR00QI01PV*`, `FR-POH_-TR00QI02PV*`, `FR-PSP-TR00QI01*`, `FR-PTW_TR01QI01PV*` | 
Turbidity     | `FR-PTW_TR01QI02PV*`, `FR-PSP-TR00QI02*`, `FR-PNB_TR00QI02PV*`, `FR-POH_-TR00QI01PV*` | 
Other         | (None of the above) | Status values of pumps, reservoirs, valves, etc

```{r, echo=FALSE}
files.vitnor <- list.files(path=dir.data, pattern="vitnor.*csv")
files.temperature <- list.files(path=dir.data, pattern="TM.*csv$")
files.flow <- list.files(path=dir.data, pattern="(FT|VO).*csv$")
files.pressure <- list.files(path=dir.data, pattern="(PT|DO)[0123456789][0123456789].*csv$")
files.conductivity <- list.files(path=dir.data, pattern="(FR-PNB_TR00QI03PV|FR-POH_-TR00QI03PV).*csv$")
files.acidity <- list.files(path=dir.data, pattern="(FR-PNB_TR00QI01PV|FR-POH_-TR00QI02PV|FR-PSP-TR00QI01|FR-PTW_TR01QI01PV).*csv$")
files.turbidity <- list.files(path=dir.data, pattern="(FR-PTW_TR01QI02PV|FR-PSP-TR00QI02|FR-PNB_TR00QI02PV|FR-POH_-TR00QI01PV).*csv$")
files.other <- Reduce(setdiff, list(files.all, files.vitnor, files.temperature, files.flow, files.pressure, files.conductivity, files.acidity, files.turbidity))
```


### Predictor and response variables

We are interested in Eventlab measurements (variable names `*vitnor*`) as **response variables**

* There are `r length(files.vitnor)` variables for Eventlab measurements
* These represent `r length(files.vitnor)/3` Eventlab sensors, each of which produce a set of three variables (`*vitnor1*`, `*vitnor2*` and `*vitnor3*`)
* Specifically, we are interested in events with values above a threshold for these variables (>1).

That leaves `r length(files.all)-length(files.vitnor)` variables that we can use as **potential predictors**:

* `r length(files.temperature)` for temperature
* `r length(files.flow)` for flow
* `r length(files.pressure)` for pressure
* `r length(files.conductivity)` for conductivity
* `r length(files.acidity)` for acidity
* `r length(files.turbidity)` for turbidity
* `r length(files.other)` other

The data quality of these variables varies. Some files do not even contain any data points.



## Exploratory Analysis

### Selection of Data Subset

```{r, echo=FALSE}
pattern.response="FR-MOBMS-vitnor1-meetwaarde"
date.begin=ymd("2015-06-25")
date.end=ymd("2015-07-05")
files.response <- list.files(path=dir.data, pattern=pattern.response)
vars.response <- createvariablefromfilename(files.response)
```

Based on large blocks of consecutive data seen in exploratory plots and multiple peaks of Eventlab variables exceeding the threshold we decided to focus on the `r vars.response` variable as response variable, in time range `r date.begin` to `r date.end`:

```{r, echo=FALSE}
setwd(dir.data)
data.response <- selecttimerange(readfiles(files.response), date.begin, date.end)
summary(data.response)
ggplot(data.response, aes_string(x="Time", y=names(data.response)[2])) + geom_point() + ggtitle("Eventlab measurements (unaggregated)")
```


### Correlations

(pair plots)

(multiple variables in single plot)

## Data Source

Raw data was provided by Vitens.