---
title: "Correlating water quality fluctuations with sensor data"
subtitle: "Vitens Open Data Challenge: Smart Water Grid"
author: "Dennis van den Berg"
date: "21/01/2016"
output: html_document
---


# Introduction

Central question: What are the correlations between changes in water quality measued by Eventlab sensors and other real-time measurements, statuses and alarm values?

See documentation: https://github.com/dljvandenberg/smart_water_grid/blob/master/doc/Vitens%20Data%20Challenge%2007122015.pdf



# Data Import

```{r, message=FALSE}
library(dplyr)
library(reshape2)
library(lubridate)
library(ggplot2)
library(caret)

dir.data <- "~/git/smart_water_grid/data"
setwd(dir.data)

readfiles <- function(fileslist) {
    listofdataframes <- lapply(fileslist, function(file) read.csv(file, 
                                                                  comment.char = "#", 
                                                                  col.names = c("Time", file), 
                                                                  colClasses = c("character", "numeric"),
                                                                  na.strings = c("", " ", "CalcFailed", "Calc Failed", "BadInput", "Bad Input", "PtCreated", "Pt Created")))

    merged <- Reduce(function(x, y) merge(x, y, all=TRUE), listofdataframes)
    
    # Convert to POSIXct/POSIXt time format
    merged$Time <- ymd_hms(merged$Time)
    
    return(merged)
}

selecttimerange <- function(dataframe, begintime = -Inf, endtime = Inf) {
    subset(dataframe, Time >= begintime & Time <= endtime)
}
```

# Data Overview

### Data files

```{r}
files.all <- list.files(path=dir.data, pattern="csv$")
```

* Our data set contains `r length(files.all)` files, all of which csv files with Time-Value pairs.
* Furthermore the data files contain some additional information in the header comments.
* The file name contains the variable (code) that was measured
    * Example: Dump_PE-FR-Deelbalansgebied-Westeinde-levering.csv
* The total size of the data set is a little over 5 GB


### Time range

* According to the documentation measurements are done in the time range January 2014 to December 2015. We confirmed this by looking at the data
* We also observe that some variables are only available in specific time ranges, often data is missing in large blocks of this 2-year time range. See example below

```{r, message=FALSE}
setwd(dir.data)
files.select <- list.files(path=dir.data, pattern="vitnor.*csv$")[1]
data.select <- readfiles(files.select)
summary(data.select)
ggplot(data.select, aes_string(x="Time", y=names(data.select)[2])) + geom_point()
```


### Variable overview

In the documentation we see the following explanation of variable names:

* Vitnor 1 / Vitnor 2 / Vitnor 3: output of Eventlab sensor (3 per sensor)
    * In case of 1 or more vitnors >1 => alarm code Orange is given. In case of 1 or more vitnors >1.5 => alarm code Red.
* Temperature: code contains 'TM'
* Flow: code contains 'FT' or 'VO' (negative values allowed in case of opposite flow direction)
* Pressure: code contains 'PT' or 'DO' (we found that these had to be followed by 2 digits in order not to select incorrect variables)
* Conductivity: 'FR-PNB_TR00QI03PV-value' or 'FR-POH_-TR00QI03PV'-value
* Acidity (pH): FR-PNB_TR00QI01PV-value, FR-POH_-TR00QI02PV-value, FR-PSP-TR00QI01-value or FR-PTW_TR01QI01PV-value
* Turbidity: FR-PTW_TR01QI02PV-value, FR-PSP-TR00QI02-value, FR-PNB_TR00QI02PV-value or FR-POH_-TR00QI01PV-value
* Miscellaneous: status values of pumps, reservoirs, valves, etc

```{r}
files.vitnor <- list.files(path=dir.data, pattern="vitnor.*csv$")
files.temperature <- list.files(path=dir.data, pattern="TM.*csv$")
files.flow <- list.files(path=dir.data, pattern="(FT|VO).*csv$")
files.pressure <- list.files(path=dir.data, pattern="(PT|DO)[0123456789][0123456789]-.*csv$")
files.conductivity <- list.files(path=dir.data, pattern="(FR-PNB_TR00QI03PV|FR-POH_-TR00QI03PV).*csv$")
files.acidity <- list.files(path=dir.data, pattern="(FR-PNB_TR00QI01PV|FR-POH_-TR00QI02PV|FR-PSP-TR00QI01|FR-PTW_TR01QI01PV).*csv$")
files.turbidity <- list.files(path=dir.data, pattern="(FR-PTW_TR01QI02PV|FR-PSP-TR00QI02|FR-PNB_TR00QI02PV|FR-POH_-TR00QI01PV).*csv$")
files.other <- Reduce(setdiff, list(files.all, files.vitnor, files.temperature, files.flow, files.pressure, files.conductivity, files.acidity, files.turbidity))
```


### Predictor and response variables

Response variables:

* We are interested in vitnor data as response variables. Specifically, we focus on values >1 for these variables
* There are `r length(files.vitnor)` variables for Eventlab measurements
* These represent `r length(files.vitnor)/3` Eventlab sensors, each of which produce a set of three variables ('vitnor1', 'vitnor2' and 'vitnor3')

Predictor variables:

* That leaves `r length(files.all)-length(files.vitnor)` variables that we can use as potential predictors
    * `r length(files.temperature)` for temperature
    * `r length(files.flow)` for flow
    * `r length(files.pressure)` for pressure
    * `r length(files.conductivity)` for conductivity
    * `r length(files.acidity)` for acidity
    * `r length(files.turbidity)` for turbidity
    * `r length(files.other)` other
* The data quality of these varies. Some files do not even contain any data points.



# Analysis



# Data Source

Raw data was provided by Vitens.